---
title: "Geostatistics in R"
author: "Ellie White"
date: "December 9th, 2016"
output:
  html_document:
    df_print: paged
---
This tutorial was written for UC Davis course HYD 273: Introduction to Geostatistics: Statistical treatment of spatial data with hydrologic emphasis. Topics: theory of random functions, variogram analysis, Kriging/co-Kriging, indicator geostatistics, and stochastic simulation of spatial variability. Geostatistical software use.  

Disclaimer: It will most definitely contain mistakes. Please let me know them at: white.elaheh@gmail.com   
Credit: Hijmans, R. (2016) Interpolation [Source code] http://www.rspatial.org for cleaned up example data and starter code.  

# Contents   
1.0 Spatial Data   
2.0 Blank Interpolation Zone   
3.0 Evaluation Metrics   
4.0 The Average Model   
5.0 Inverse Distance Weighted Model   
6.0 Variogram   
    .... 6.1 Anisotropy   
7.0 Kriging Models   
    .... 7.1 Simple Kriging   
    .... 7.2 Ordinary Kriging   
    .... 7.3 Universal Kriging   
    .... 7.4 Indicator Kriging    
    .... 7.5 Local Kriging   
    .... 7.6 CoKriging   
    .... 7.7 Block Kriging   
8.0 Sequential Gaussian Simulation   
9.0 Comparing Models  
    .... 9.1 Comparing RMSE   
    .... 9.2 Comparing Plots   
10.0 Ensemble Model  
11.0 Transition Probability Markov Chain   

# Libraries  
library(sp)     --for geostatistical analysis   
library(rgdal)  --for spatial data transformations   
library(raster) --for spatial data manipulation and analysis   
library(gstat)  --for geostatistical analysis   
library(dismo)  --for k fold cross-validation  
library(spMC)   --for transition probability markov chains  

**note:** install each of these libraries by typing the following in your Console: install.packages("[name of library]")  

```{r setup, include=FALSE}
library(knitr)
library(formatR)
opts_chunk$set(
  fig.width =7.5,
  fig.height=7.5,
  collapse  =TRUE,
  tidy      =TRUE, 
  tidy.opts=list(width.cutoff=120)
)
```

# 1.0 Spatial Data  
**What:** Ozone Concentrations    
**Units:** ppb (parts per billion)  
**Time:** averages for 1980-2009  
**Time resolution:** 30 yr average  
**Data type:** .csv  
**Data Source:** https://www.arb.ca.gov/aqd/aqdcd/aqdcddld.htm, or download the clean version from: http://rspatial.org/analysis/data/airqual.csv
**Provided by:** Cal EPA Air Resources Board

```{r data}
# download the clean version from: http://rspatial.org/analysis/data/airqual.csv, read in the data from your working directory  
aq_data <- read.csv("inputdata/airqual.csv")
head(aq_data)

# change the OZDLYAV column to ppm (parts per million) so it is easier to read. This is the column of data we will be working with  
aq_data$OZDLYAV <- aq_data$OZDLYAV*1000
aqd <- aq_data 

# change data frame to a spatial points data frame
library(sp)
coordinates(aqd) <- ~LONGITUDE + LATITUDE
proj4string(aqd) <- CRS('+proj=longlat +datum=NAD83')

# transform data into a suitable projection, for California: Albers and Teale Albers are good
TA <- CRS("+proj=aea +lat_1=34 +lat_2=40.5 +lat_0=0 +lon_0=-120 +x_0=0 +y_0=-4000000 +datum=NAD83 +units=km +ellps=GRS80")

library(rgdal)
aq_ta <- spTransform(aqd, TA)

# for visualization purposes download California counties from link in: http://rspatial.org/analysis/rst/4-interpolation.html 
library(raster)
ca <- shapefile("Counties/counties_2000.shp")
ca_ta <- spTransform(ca, TA)

# visualization of the data
plot(ca_ta, border='gray', axes=TRUE, main="Ozone Concentration (ppm)")
text(aq_ta, labels=round(aq_ta$OZDLYAV, 2), cex=.5, col='navy')

ucd_colors <- colorRampPalette(c("goldenrod", "navy"))
spplot(aq_ta, "OZDLYAV", col.regions=ucd_colors(6), key.space="right", main="Ozone Concentration (ppm)", sp.layout=list("sp.polygons", ca_ta, col="grey"))

hist(aq_data$OZDLYAV, main="Histogram", col="navy", border="goldenrod", xlab="OZDLYAV (ppm)")

hist(log(aq_data$OZDLYAV), main="Histogram", col="navy", border="goldenrod", xlab="log(OZDLYAV)")
```

# 2.0 Blank Interpolation Zone
This section creates a spatial grid over california (projected to Teale Albers) to interpolate or make predictions to, i.e. a blank slate to be later given values with each model developed.  
```{r interpolation_zone}
r <- raster(ca_ta)
res(r) <- 10  # 10 km if your Coordinate Reference System's units are in km, the Teal Albers system is in km

# coerce r into a spatial grid, not very different from a raster
g <- as(r, 'SpatialGrid')
```

# 3.0 Evaluation Metrics  
This section defines RMSE, R Squared, and NSE as functions to be used when evaluating and comparing models.  
```{r rmse}
RMSE <- function(observed, predicted) {
  sqrt(mean((predicted - observed)^2, na.rm=TRUE))
}

R2 <- function(observed, predicted) {
  sum((predicted - mean(observed, na.rm=TRUE))^2, na.rm=TRUE)/sum((observed - mean(observed, na.rm=TRUE))^2, na.rm=TRUE)
}

# use this if you have hydrologic data, I have left it out of this analysis
NSE <- function(observed, predicted) {
  1-(sum((predicted - observed)^2, na.rm=TRUE)/sum((mean(observed) - observed)^2, na.rm=TRUE))
}
```

# 4.0 The Average Model  
This model is the arithmatic average of all the observations.  
```{r average_model}
am <- mean(aqd$OZDLYAV)

# make a raster that has this value everywhere, we may need this for comparison to other models later.
am_raster <- r
values(am_raster) <- am

# evaluation of this model
am_rmse <- RMSE(am, aq_ta$OZDLYAV)
am_rmse
```

# 5.0 Inverse Distance Weighted Model  
This model is a weighted combination of all observations. The weights will be purely based on distance.  
```{r idw}
library(gstat)
# gstat objects hold all the information necessary for univariate or multivariate geostatistical prediction
gs <- gstat(formula=OZDLYAV~1, locations=aq_ta)

idw <- interpolate(r, gs)
idwr <- mask(idw, ca_ta)

# plot
plot(idwr, main="IDW")
plot(ca_ta, border='gray', add=TRUE)
contour(idwr, add=TRUE, nlevels=10)
```

## 5.1 Optimal Inverse Distance Weighted Model  
In the IDW model there are two parameters that can be optimized: 1) the number of nearest observations that should be used for a prediction, and 2) the distance decay rate.  
```{r opt_idw}
# optimization objective: minimize the RMSE
f1 <- function(x, test, train) {
  nmx <- x[1]
  idp <- x[2]
  if (nmx < 1) return(Inf)
  if (idp < .001) return(Inf)
  m <- gstat(formula=OZDLYAV~1, locations=train, nmax=nmx, set=list(idp=idp))
  p <- predict(m, newdata=test, debug.level=0)$var1.pred
  RMSE(test$OZDLYAV, p)
}

# 5-fold cross validation
set.seed(20161209)
i <- sample(nrow(aq_ta), 0.2*nrow(aq_ta))
tst <- aq_ta[i,]
trn <- aq_ta[-i,]

# optimization model
opt <- optim(c(8, .5), f1, test=tst, train=trn)
opt

# optimal IDW model  
# opt$par[1]: the number of nearest observations that should be used for a prediction
# opt$par[2]: the distance decay rate. the power on the distance in the demoninator
idw_opt <- gstat(formula=OZDLYAV~1, locations=aq_ta, nmax=opt$par[1], set=list(idp=opt$par[2]))
idw_b <- interpolate(r, idw_opt)
idw_best <- mask(idw_b, ca_ta)

# plot
plot(idw_best, main="IDW_BEST")
plot(ca_ta, border='gray', add=TRUE)
contour(idw_best, add=TRUE, nlevels=10)
```

# 6.0 Variogram  
A variogram is a description of the spatial continuity of the data. The experimental variogram is a function fitted to the variogram. This function measures the variability between pairs of points at various distances and is used in Kriging models.  
```{r variogram}
gs <- gstat(formula=OZDLYAV~1, locations=aq_ta) 
v <- variogram(gs, width=20) # width is the width of the bins for grouping data
head(v)
plot(v, main="Variogram (Bins=20 km)")

# set cloud=TRUE to see all the pairs of observations
vc <- variogram(gs, width=20, cloud=TRUE)
head(vc)
plot(vc, main="Variogram Cloud")

# now to fit a model to the variogram. Here are all the types of fits we can try
show.vgms()

# fit a variogram model, best way is to eye ball the sill, range, and nugget
# here is an exponential fit
fve <- fit.variogram(v, vgm(85, "Exp", 75, 20))
fve
plot(variogramLine(fve, 400), type='l', ylim=c(0,120), main="Exponential Fit")
points(v[,2:3], pch=20, col='navy', cex=1.5)

# here is a spherical fit
fvs <- fit.variogram(v, vgm(85, "Sph", 75, 20))
fvs
plot(variogramLine(fvs, 400), type='l', ylim=c(0,120), main="Spherical Fit")
points(v[,2:3], pch=20, col='navy', cex=1.5)
```

## 6.1 Anisotropy  
Anisotropy is when the range (geometric anisotropy) or the sill (zonal anisotropy) of the semivariogram varies in different directions. Both types of anisotropy can be present.  
```{r anisotropy}
# use alpha to set the direction
vgm_aniso <- variogram(gs, alpha=c(0, 45, 90, 135))
plot(vgm_aniso)

# eye ball a model fit, eyeball the sill, range, and nugget
aniso_fit <- vgm(200, "Sph", 200, 0, anis=c(90, 1/3))
plot(vgm_aniso, aniso_fit, as.table=TRUE)

# that wasn't very good. let R try to fit a model through regression
raniso_fit <- fit.variogram(vgm_aniso, model=aniso_fit)
plot(vgm_aniso, raniso_fit)
 
# can also subset and eyeball a model fit in the 45 Degree direction
vgm_45 <- subset(vgm_aniso, vgm_aniso$dir.hor == 45)  
plot(vgm_45)
vgm_45_sph <- vgm(150, "Sph", 200, 0)
plot(vgm_45, vgm_45_sph, main="Variogram (Dir=45 Degrees)") # notice this is another way of plotting variograms
```

# 7.0 Kriging Models
Originating in geostatistics, kriging or Gaussian process regression is a method of interpolation for which the interpolated values are modeled by a Gaussian process governed by prior covariances. We will look at a few different types of kriging below.  

## 7.1 Simple Kriging --known mean
For simple kriging (and simulation based on simple kriging) you need to define a vector with the trend coefficients (including intercept); if no independent variables are defined the model only contains an intercept and beta should be the simple kriging mean.  
```{r sk}
skp <- krige(formula=OZDLYAV~1, locations=aq_ta, g, model=fve, beta=mean(aq_data$OZDLYAV))
spplot(skp)

# plot nicely
sk <- brick(skp)
sk <- mask(sk, ca_ta)
names(sk) <- c('prediction', 'variance')
plot(sk)
title(main="Simple Kriging", line=6)

plot(sk$prediction, main="Simple Kriging")
plot(ca_ta, border='gray', add=TRUE)
contour(sk$prediction, add=TRUE, nlevels=10)
```

## 7.2 Ordinary Kriging --no known mean
Leave beta out of simple kriging to get an ordinary kriging estimate.  
```{r ok}
okp <- krige(formula=OZDLYAV~1, locations=aq_ta, g, model=fve)
spplot(okp)

# plot nicely
ok <- brick(okp)
ok <- mask(ok, ca_ta)
names(ok) <- c('prediction', 'variance')
plot(ok)
title(main="Ordinary Kriging", line=6)

plot(ok$prediction, main="Ordinary Kriging")
plot(ca_ta, border='gray', add=TRUE)
contour(ok$prediction, add=TRUE, nlevels=10)
```

## 7.3 Universal Kriging --kriging with external drift  
Universal kriging is useful when the variable to map is related to other spatially known variables. For universal kriging, suppose z is a trending surface either linearly or quadratically dependent on x and y. This section shows you how to do both.    
```{r uk}
# set up the x and y coordinates 
aq_ta$x <- coordinates(aq_ta)[, 1]
aq_ta$y <- coordinates(aq_ta)[, 2]

# cannot use g, a spatial grid, anymore. The grid here needs to be a spatial pixels data frame! 
gpix <- as(r, 'SpatialPixelsDataFrame')

# linear trend surface
ukpl <- krige(formula=OZDLYAV~x+y, locations=aq_ta, gpix, model=fve)
spplot(ukpl)

# plot nicely
ukl <- brick(ukpl)
ukl <- mask(ukl, ca_ta)
names(ukl) <- c('prediction', 'variance')
plot(ukl)
title(main="Universal Kriging-Linear Trend Surface", line=6)

plot(ukl$prediction, main="Universal Kriging-Linear Trend Surface")
plot(ca_ta, border='gray', add=TRUE)
contour(ukl$prediction, add=TRUE, nlevels=10)

# quadratic trend surface 
ukpq <- krige(formula=OZDLYAV~ x + y + I(x*y) + I(x^2) + I(y^2), locations=aq_ta, gpix, model=fve)
spplot(ukpq)

# plot nicely
ukq <- brick(ukpq)
ukq <- mask(ukq, ca_ta)
names(ukq) <- c('prediction', 'variance')
plot(ukq)
title(main="Universal Kriging-Quadratic Trend Surface", line=6)

plot(ukq$prediction, main="Universal Kriging-Quadratic Trend Surface")
plot(ca_ta, border='gray', add=TRUE)
contour(ukq$prediction, add=TRUE, nlevels=10)
```

## 7.4 Indicator Kriging   
Use indicator kriging to map the probability of ozone exceeding a critical threshold.    
```{r ik}
# pick a threshold, lets look at our variable again
summary(aq_data$OZDLYAV)

# develop a variogram for indicator kriging 
gsi <- gstat(formula=I(OZDLYAV>30)~1, locations=aq_ta) 
vi <- variogram(gsi, width=20)
fvis <- fit.variogram(vi, vgm(85, "Exp", 75, 20))

plot(variogramLine(fvis, 400), type='l', ylim=c(0,0.3), main="Exponential Fit")
points(vi[,2:3], pch=20, col='navy', cex=1.5)

# indicator kriging
ikp <-  krige(formula=I(OZDLYAV>30)~1, locations=aq_ta, g, model=fvis, nmax=30, indicators=TRUE)

spplot(ikp)

# plot nicely
ik <- brick(ikp)
ik <- mask(ik, ca_ta)
names(ik) <- c('prediction', 'variance')
plot(ik)
title(main="Indicator Kriging (i>30ppm)", line=6)
```

## 7.5 Local Kriging  
In kriging, by default, all observations are used. But we can use nmin, nmax and or maxdist to restrict the number of observations the kriging estimator can use.  
**nmax:** the max number of nearest observations that should be used for a kriging prediction or simulation, where nearest is defined in terms of the space of the spatial locations.  
**maxdist:** only observations within a distance of maxdist from the prediction location are used for prediction or simulation; if combined with nmax, both criteria apply.  
**nmin:** if the number of nearest observations within distance maxdist is less than nmin, a missing value will be generated; see maxdist.  
```{r lk}
# remember the dimensions of the data 
dim(aq_data) 

# and the clustering of data points
plot(ca_ta, border='gray', main="Data Points")
points(aq_ta, col='navy')

# now pick an appropriate nmin, nmax, and or maxdist
lkp <- krige(formula=OZDLYAV~1, locations=aq_ta, g, model=fve, nmin=20, nmax=100, maxdist=100)
spplot(lkp)

# plot nicely
lk <- brick(lkp)
lk <- mask(lk, ca_ta)
names(lk) <- c('prediction', 'variance')
plot(lk)
title(main="Local Kriging", line=6)

plot(lk$prediction, main="Local Kriging")
plot(ca_ta, border='gray', add=TRUE)
contour(lk$prediction, add=TRUE, nlevels=10)
```

## 7.6 CoKriging  
Use cokriging, or making use of other data sets, to make a better prediction. If the new data is grounded in pysicall processes that influence the variable of interest, the variance in the error prediction can be reduced.  
```{r cok}
# find potentially useful data sets, use scatterplot matrix

# this function will just print the r and pvalue in our scatterplot matrix below
panel.cor <- function(x, y, digits=2, cex.cor, ...){
  usr <- par("usr"); on.exit(par(usr))
  par(usr=c(0, 1, 0, 1))
  # correlation coefficient
  r <- cor(x, y)
  txt <- format(c(r, 0.123456789), digits=digits)[1]
  txt <- paste("r= ", txt, sep="")
  text(0.5, 0.6, txt)

  # p-value calculation
  p <- cor.test(x, y)$p.value
  txt2 <- format(c(p, 0.123456789), digits=digits)[1]
  txt2 <- paste("p= ", txt2, sep="")
  if(p<0.01) txt2 <- paste("p= ", "<0.01", sep="")
  text(0.5, 0.4, txt2)
}

pairs(aq_data[,22:28], upper.panel=panel.cor)

# based on the r values in the scatter plot matrix above, I will be using the following data sets
summary(aq_ta$OZMAX8O)
summary(aq_ta$OZMX8ST)

# now to create a gstat object and the variogram model
gsc1 <- gstat(id="OZDLYAV", formula=OZDLYAV~1, locations=aq_ta) 
gsc2 <- gstat(gsc1, id="OZMAX8O", formula=OZMAX8O~1, locations=aq_ta) 
gsc3 <- gstat(gsc2, id="OZMX8ST", formula=OZMX8ST~1, locations=aq_ta) 

# plot the variograms and cross-variograms
plot(variogram(gsc3))

# fit a model variogram to the target variogram
fvsc <- fit.variogram(variogram(gsc1), vgm(85, "Sph", 75, 20)) 

# fit a model variogram to all the variograms
vmc <- variogram(gsc3) 
fvmc <- fit.lmc(vmc, gsc3, model=fvsc) 

# plot the fitted variograms to all the sample variograms:
plot(variogram(gsc3), fvmc)

# co-kriging, WARNING: this may take a few minutes!
ckp <- predict(fvmc, g) 
spplot(ckp)

# plot nicely
ck <- brick(ckp)
ck <- mask(ck, ca_ta)
plot(ck)

plot(ck$OZDLYAV.pred, main="Co-Kriging")
plot(ca_ta, border='gray', add=TRUE)
contour(ck$OZDLYAV.pred, add=TRUE, nlevels=10)
```

## 7.7 Block Kriging --for rectangular or irregular blocks  
Use block kriging to smooth out the surface, and have a less computationally expensive model.  
**block size:** a vector with 1, 2 or 3 values containing the size of a rectangular in x-, y- and z-dimension respectively (0 if not set), or a data frame with 1, 2 or 3 columns, containing the points that discretize the block in the x-, y- and z-dimension to define irregular blocks relative to (0,0) or (0,0,0)—see also the details section of predict.gstat.  
```{r bk}
# let's try smaller blocks first
bkp_small <- krige(formula=OZDLYAV~1, locations=aq_ta, g, model=fve, block=c(50, 50))
spplot(bkp_small)

# plot nicely
bk_small <- brick(bkp_small)
bk_small <- mask(bk_small, ca_ta)
names(bk_small) <- c('prediction', 'variance')
plot(bk_small)
title(main="Block Kriging (b=50km)", line=6)

plot(bk_small$prediction, main="Block Kriging (b=50km)")
plot(ca_ta, border='gray', add=TRUE)
contour(bk_small$prediction, add=TRUE, nlevels=10)

# let's try larger blocks
bkp_large <- krige(formula=OZDLYAV~1, locations=aq_ta, g, model=fve, block=c(400, 400))
spplot(bkp_large)

# plot nicely
bk_large <- brick(bkp_large)
bk_large <- mask(bk_large, ca_ta)
names(bk_large) <- c('prediction', 'variance')
plot(bk_large)
title(main="Block Kriging (b=400km)", line=6)

plot(bk_large$prediction, main="Block Kriging (b=400km)")
plot(ca_ta, border='gray', add=TRUE)
contour(bk_large$prediction, add=TRUE, nlevels=10)
```

# 8.0 Sequential Gaussian Simulation  
Sequential Gaussian simulation (SGS) simulates continuous variables, such as our ozone concentration data. Sequential indicator simulation (SIS) simulates discrete variables, such as an ozone indicator variable that splits the ozone concentration to high and low values based on a threshold value.  
**nsim:** integer; if set to a non-zero value, conditional simulation is used instead of kriging interpolation. For this, sequential Gaussian or indicator simulation is used (depending on the value of indicators), following a single random path through the data.  
```{r conditional_sim}
# SGS
condsimp <-  krige(formula=OZDLYAV~1, locations=aq_ta, g, model=fve, nmax=30, nsim=4)
spplot(condsimp, main="four conditional simulations")

# plot nicely
condsim <- brick(condsimp)
condsim <- mask(condsim, ca_ta)
plot(condsim)

# SIS, note that the model should be set to the fit to the indicator variogram!
condsimp2 <-  krige(formula=I(OZDLYAV>30)~1, locations=aq_ta, g, model=fvis, nmax=30, nsim=4, indicators=TRUE)
spplot(condsimp2, main="four conditional simulations")

# plot nicely
condsim2 <- brick(condsimp2)
condsim2 <- mask(condsim2, ca_ta)
plot(condsim2)
```

# 9.0 Comparing Models  
To recap, the models developed are: Average Model(am), Inverse Distance Weighted (idw), Simple Kriging (sk), Ordinary Kriging (ok), Universal Kriging with a Linear Trend (ukl), Universal Kriging with a Quadratic Trend (ukq), Indicator Kriging (ik), Co-Kriging (ck), Block Kriging (bk). 

note: I left co-kriging out of the cross validation, because the code would take a looooong time to run.  

## 9.1 Comparing RMSE and NSE 
Use k-fold crossvalidation to evaluate the models.   

```{r compare_rmse}
library(dismo)

nfolds <- 5
k <- kfold(aq_ta, nfolds)

# define vectors to be populated later
ensrmse <- idwrmse <- skrmse <- okrmse <- uklrmse <- ukqrmse <- ikrmse <- lkrmse <- bk_lrmse <- rep(NA, 5)
ensrsq <- idwrsq <- skrsq <- okrsq <- uklrsq <- ukqrsq <- ikrsq <- lkrsq <- bk_lrsq <- rep(NA, 5)

# cross validation loop
for (i in 1:nfolds) {
  test <- aq_ta[k!=i,]
  train <- aq_ta[k==i,]
  
  # idw_best
  m <- gstat(formula=OZDLYAV~1, locations=train, nmax=opt$par[1], set=list(idp=opt$par[2]))
  p1 <- predict(m, newdata=test, debug.level=0)$var1.pred
  idwrmse[i] <- RMSE(test$OZDLYAV, p1)
  idwrsq[i] <- R2(test$OZDLYAV, p1)

  # sk
  m <- gstat(formula=OZDLYAV~1, locations=train, model=fve, beta=mean(aq_ta$OZDLYAV))
  p2 <- predict(m, newdata=test, debug.level=0)$var1.pred
  skrmse[i] <- RMSE(test$OZDLYAV, p2)
  skrsq[i] <- R2(test$OZDLYAV, p2)
  
  # ok
  m <- gstat(formula=OZDLYAV~1, locations=train, model=fve)
  p3 <- predict(m, newdata=test, debug.level=0)$var1.pred
  okrmse[i] <- RMSE(test$OZDLYAV, p3)
  okrsq[i] <- R2(test$OZDLYAV, p3)
  
  # ukl
  m <- gstat(formula=OZDLYAV~x+y, locations=aq_ta, model=fve)
  p4 <- predict(m, newdata=test, debug.level=0)$var1.pred
  uklrmse[i] <- RMSE(test$OZDLYAV, p4)
  uklrsq[i] <- R2(test$OZDLYAV, p4)
  
  # ukq
  m <- gstat(formula=OZDLYAV~x+y+I(x*y)+I(x^2)+I(y^2), locations=aq_ta, model=fve)
  p5 <- predict(m, newdata=test, debug.level=0)$var1.pred
  ukqrmse[i] <- RMSE(test$OZDLYAV, p5)
  ukqrsq[i] <- R2(test$OZDLYAV, p5)
  
  # ik
  m <- gstat(formula=I(OZDLYAV>30)~1, locations=aq_ta, model=fvis, nmax=30)
  p6 <- predict(m, newdata=test, debug.level=0, indicators=TRUE)$var1.pred
  test$indicator <- ifelse(test$OZDLYAV>30, 1, 0)
  ikrmse[i] <- RMSE(test$indicator, p6)
  ikrsq[i] <- R2(test$indicator, p6)
  
  # lk
  m <- gstat(formula=OZDLYAV~1, locations=aq_ta, model=fve, nmin=20, nmax=100, maxdist=100)
  p7 <- predict(m, newdata=test, debug.level=0)$var1.pred
  lkrmse[i] <- RMSE(test$OZDLYAV, p7)
  lkrsq[i] <- R2(test$OZDLYAV, p7)
  
  # bk
  m <- gstat(formula=OZDLYAV~1, locations=aq_ta, model=fve)
  p8 <- predict(m, newdata=test, debug.level=0, block=c(400, 400))$var1.pred
  bk_lrmse[i] <- RMSE(test$OZDLYAV, p8)
  bk_lrsq[i] <- R2(test$OZDLYAV, p8)
  
  w <- c(idwrsq[i], skrsq[i], okrsq[i], uklrsq[i], ukqrsq[i], ikrsq[i], lkrsq[i], bk_lrsq[i])
  weights <- w / sum(w)
  
  # ensemble model
  ensemble <- p1*weights[1]+ p2*weights[2]+ p3*weights[3]+ p4*weights[4]+ p5*weights[5]+ p6*weights[6]+ p7*weights[7]+ p8*weights[8]
  ensrmse[i] <-  RMSE(test$OZDLYAV, ensemble)
  ensrsq[i] <- R2(test$OZDLYAV, ensemble)
}

idw_rmse <- mean(idwrmse)
sk_rmse <- mean(skrmse)
ok_rmse <- mean(okrmse) 
ukl_rmse <- mean(uklrmse) 
ukq_rmse <- mean(ukqrmse) 
ik_rmse <- mean(ikrmse)
lk_rmse <- mean(lkrmse) 
bk_lrmse <- mean(bk_lrmse)

idw_rsq <- mean(idwrsq)
sk_rsq <- mean(skrsq)
ok_rsq <- mean(okrsq) 
ukl_rsq <- mean(uklrsq) 
ukq_rsq <- mean(ukqrsq) 
ik_rsq <- mean(ikrsq)
lk_rsq <- mean(lkrsq) 
bk_lrsq <- mean(bk_lrsq)

rms <- c(idw_rmse, sk_rmse, ok_rmse, ukl_rmse, ukq_rmse, ik_rmse, lk_rmse, bk_lrmse)
rsq <- c(idw_rsq, sk_rsq, ok_rsq, ukl_rsq, ukq_rsq, ik_rsq, lk_rsq, bk_lrsq)
names(rms) <- names(rsq) <- c('IDW_Best', 'SK', 'OK', 'UK_Lin', 'UK_Quad', 'IK', 'LK', 'BK_Large')
rms
rsq

# barplot the RMSE, store and print the rmse on the bars
ylim <- c(0, 1.1*max(rms))
xx <- barplot(rms, main="RMSE Comparison", xlab="Geostatistical Methods", ylim=ylim, col="goldenrod")
text(x=xx, y=rms, label=round(rms, digits=2), pos=3, cex=0.8, col="navy")

# barplot the RSQ, store and print the rmse on the bars
ylim <- c(0, 1.1*max(rsq))
xx <- barplot(rsq, main="R^2 Comparison", xlab="Geostatistical Methods", ylim=ylim, col="goldenrod")
text(x=xx, y=rsq, label=round(rsq, digits=2), pos=3, cex=0.8, col="navy")
```

## 9.2 Comparing Plots  
Use plots to identify spatial differences and maybe go back to tweek model parameters.   
```{r comparing_plots}
# compare prediction plots
s <- stack(idw_best, sk[[1]], ok[[1]], ukl[[1]], ukq[[1]], ik[[1]], lk[[1]], ck[[1]], bk_large[[1]])
names(s) <- c('IDW_Best', 'SK', 'OK', 'UK_Lin', 'UK_Quad', 'IK', 'LK', 'Co_K', 'BK_Large')
plot(s)

# compare errors plots
s_stdev <- stack(sk[[2]], ok[[2]], ukl[[2]], ukq[[2]], ik[[2]], lk[[2]], ck[[2]], bk_large[[2]])
names(s_stdev) <- c('SK_error', 'OK_error', 'UK_Lin_error', 'UK_Quad_error', 'IK_error', 'LK_error', 'Co_K_error', 'BK_Large_error')
plot(s_stdev)
```

# 10.0 Ensemble Model  
We can combine any or all models by taking a weighted average of all the model predictions with R^2 as the weights.  
```{r ensemble}
# from cross validation we have:
rmse_e <- mean(ensrmse)
rmse_e

rsq_e <- mean(ensrsq)
rsq_e

# weights were calculated in the cv loop
names(weights) <- c('IDW_Best', 'SK', 'OK', 'UK_Lin', 'UK_Quad', 'IK', 'LK', 'BK_Large')
weights

models <- stack(idw_best, sk[[1]], ok[[1]], ukl[[1]], ukq[[1]], ik[[1]], lk[[1]], bk_large[[1]])
ensemble_model <- sum(models*weights)

plot(ensemble_model, main="Ensemble Model")
plot(ca_ta, border='gray', add=TRUE)
contour(ensemble_model, add=TRUE, nlevels=10)
```

# 11.0 Transition Probability Markov Chain  
Much like the TPROGS software, the spMC lirary is used for modelling of transition probability matrices, and implementing efficient algorithms for improving predictions and simulations of categorical random fields. It deals with both one-dimensional and multidimensional continuous lag models.  

Three different fitting methods were implemented in the package. The first is based on the estimates of the main features that characterise the process, the second focuses on the minimisation of the discrepancies between the empirical and theoretical transition probabilities, and the third follows the maximum entropy approach. Once the model parameters are properly estimated, transition probabilities are calculated through the matrixvalued exponential function.  

For the estimations of one-dimensional continuous lag models use:  
**transiogram:** Empirical transition probabilities estimation  
**tpfit:** One-dimensional model parameters estimation  
**tpfit_ils:** Iterated least squares method for one-dimensional model parameters estimation  
**tpfit_me:** Maximum entropy method for one-dimensional model parameters estimation  
**tpfit_ml:** Mean length method for one-dimensional model parameters estimation  

For the estimations of multidimensional continuous lag models use:  
**pemt:** Pseudo-empirical multidimensional transiograms estimation  
**multi_tpfit:** Multidimensional model parameters estimation  
**multi_tpfit_ils:** Iterated least squares method for multidimensional model parameters estimation  
**multi_tpfit_me:** Maximum entropy method for multidimensional model parameters estimation  
**multi_tpfit_ml:** Mean length method for multidimensional model parameters estimation  

These transition probabilities are then combined to predict the category in an unsampled position. Three algorithms are used to simulate spatial random fields; those based on the kriging techniques (Carle and Fogg, 1996), those using fixed and random path methods (Li, 2007a; Li and Zhang, 2007), or those using multinomial categorical simulation proposed by Allard et al. (2011).  
For categorical spatial random field simulation and prediction use:  
**sim:** Random field simulation  
**sim_ck:** Conditional simulation based on indicator cokriging  
**sim_ik:** Conditional simulation based on indicator kriging  
**sim_mcs:** Multinomial categorical simulation  
**sim_path:** Conditional simulation based on path algorithms  

For more information see the spMC package (Sartore, 2016).  
```{r transition_probability}
# change the ozone concentration data into categorical data 

# make the groups, you can see the interval categories now in the data set
aq_ta$group <- cut(aq_ta$OZDLYAV, breaks=6)
table(aq_ta$group)

# we can lump high and very high, because we don't have enough data, but I will leave it for now

# make the groupings into categories: verylow, low, etc.
aq_ta$categories <- cut(aq_ta$OZDLYAV, breaks=6, labels=c("verylow", "low", "medlow", "medhigh", "high", "veryhigh"))
table(aq_ta$categories)

# plot the categorical variable now
library(RColorBrewer)
brwCol <- brewer.pal(nlevels(aq_ta$categories), "YlGnBu")
spplot(aq_ta, "categories", col.regions=brwCol, key.space="right", main="Ozone Concentration Categories", sp.layout=list("sp.polygons", ca_ta, col="grey"), pch=15)

# now we can look at the spatial transition probabilities in these categories
library(spMC)

# make two transiograms to estimate two transition rate matrices, i.e. the model coefficients, along the axial directions 
# first, all couples of points along axial directions are chosen such that their lag-length is less than three km. After, we calculate the empirical transition probabilities for twenty points within the maximum distance
Trg <- list()
Trg[[1]] <- transiogram(aq_ta$categories, aq_ta@coords, max.dist=30, mpoints=50, direction=c(1, 0))
Trg[[2]] <- transiogram(aq_ta$categories, aq_ta@coords, max.dist=30, mpoints=50, direction=c(0, 1))

# if we want to compare these probabilities with the theoretical one, we first need to estimate two transition rate matrices, i.e. the model coefficients, along the axial directions

# estimate the transition rates through the mean lengths; estimate the parameters of the MC model
RTm <- list()
RTm[[1]] <- tpfit_ml(aq_ta$categories, aq_ta@coords, direction=c(1, 0))
RTm[[2]] <- tpfit_ml(aq_ta$categories, aq_ta@coords, direction=c(0, 1))
RTm

# estimate the transition rate matrices through the maximum entropy approach
ETm <- list()
ETm[[1]] <- tpfit_me(aq_ta$categories, aq_ta@coords, direction=c(1, 0))
ETm[[2]] <- tpfit_me(aq_ta$categories, aq_ta@coords, direction=c(0, 1))
ETm

# given the model coefficients, the transition probabilities for some specific lags are calculated
RTr <- list()
ETr <- list()
for (i in 1:2) {
  RTr[[i]] <- predict(RTm[[i]], lags=Trg[[i]]$lags)
  ETr[[i]] <- predict(ETm[[i]], lags=Trg[[i]]$lags)
}

# Since these probabilities are calculated with respect to some fixed directions, i.e. by considering a one-dimensional perspective, they can be graphically compared

#  a transition probability matrix plot, it shows the probability dynamic related to one-dimensional lags along the specified direction

for (i in 1:2){
  mixplot(list(Trg[[i]], RTr[[i]], ETr[[i]]), type=c("p", "l", "l"), pch="+", col=c(3, 1, 2), legend=FALSE, main=paste("One-dimensional transiograms", c("(X-axis)", "(Y-axis)")[i]))
}

# The function pemt() can be considered as another tool to check the anisotropy of the process. It estimates the transition rate matrix for each multidimensional lag direction and computes the transition probabilities. At the same time the function calculates other probabilities through the transition rates. Then the probabilities are drawn.
psEmpTr <- pemt(aq_ta$categories, aq_ta@coords, mpoints=40, max.dist=c(50, 50))

# dev.off() # uncomment and run this line if plot is not running
image(psEmpTr, col=rev(heat.colors(500)), useRaster=TRUE, breaks=c(0:500)/500, contour=FALSE,  mar=c(0.1, 0.1, 0.1, 0.1))

# plot again with contour lines that are displayed for both the pseudo-empirical and the theoretical probabilities
# dev.off()
contour.pemt(psEmpTr, mar=c(0.1, 0.1, 0.1, 0.1), col=c("goldenrod", "navy"))

# methods to estimate transition rate matrices along axial directions. These functions implement algorithms based on the mean lengths and maximum entropy
MTr <- list()
MTr$average <- multi_tpfit_ml(aq_ta$categories, aq_ta@coords)
MTr$entropy <- multi_tpfit_me(aq_ta$categories, aq_ta@coords)

# with the output of these functions, we can draw the theoretical transition probability maps
# dev.off()
image(MTr$average, 40, max.dist=50, col=rev(heat.colors(500)), nlevels=5, breaks=0:500/500)
# dev.off()
image(MTr$entropy, 40, max.dist=50, col=rev(heat.colors(500)), nlevels=5, breaks=0:500/500)

# now, the simulation 

# plot the map of observations again
X <- aq_ta$x
Y <- aq_ta$y

par(mfrow=c(1, 1), mar=c(5, 4, 4, 2))
plot(X, Y, col=brwCol[aq_ta$categories], pch=15, main="Ozone Concentration Categories")
legend("topright", legend=levels(aq_ta$categories), col=brwCol, pch=15)

# sample observations
set.seed(20161209)
smp <- sample(length(aq_ta):1, 100)

# plot the sample
plot(X, Y, type="n", main="Sample of 100 observations", xlab="X", ylab="Y")
points(X[smp], Y[smp], pch=19, col=brwCol[aq_ta$categories[smp]])
legend("topright", legend=levels(aq_ta$categories), col=brwCol, pch=19)

# make a grid
grid <- expand.grid(X,Y)

# the kriging algorithm will approximate the conditional probabilities by considering the twelve nearest neighbors for all points in the simulation grid.
iks <- sim_ik(MTr$average, data=aq_ta$categories[smp], coords=aq_ta@coords[smp,], grid, knn=12)

# Both fixed and random path simulation methods are performed by considering those nearest points along the axial directions within a radius of length ten.
fpth <- sim_path(MTr$average, data=aq_ta$categories[smp], coords=aq_ta@coords[smp,], grid, radius=1, TRUE)
rpth <- sim_path(MTr$average, data=aq_ta$categories[smp], coords=aq_ta@coords[smp,], grid, radius=1)

# The multinomial categorical simulation method will approximate the prediction probabilities by considering all sample points.
mcs <- sim_mcs(MTr$average, data=aq_ta$categories[smp], coords=aq_ta@coords[smp, ], grid)

posCol <- as.integer(iks$Prediction)
plot(X, Y, pch=15, col=brwCol[posCol], main="Kriging prediction map")
legend("topright", legend=levels(aq_ta$categories), col=brwCol, pch=15)
posCol <- as.integer(fpth$Prediction)
plot(X, Y, pch=15, col=brwCol[posCol], main="Fixed path prediction map")
legend("topright", legend=levels(aq_ta$categories), col=brwCol, pch=15)
posCol <- as.integer(rpth$Prediction)
plot(X, Y, pch=15, col=brwCol[posCol], main="Random path prediction map")
legend("topright", legend=levels(aq_ta$categories), col=brwCol, pch=15)
posCol <- as.integer(mcs$Prediction)
plot(X, Y, pch=15, col=brwCol[posCol],
main="Multinomial categorical prediction map")
legend("topright", legend=levels(aq_ta$categories), col=brwCol, pch=15)

# to compute the number of matches, we calculate the contingency table
ikTb <- table(aq_ta$categories[smp], iks$Prediction[smp])
fpTb <- table(aq_ta$categories[smp], fpth$Prediction[smp])
rpTb <- table(aq_ta$categories[smp], rpth$Prediction[smp])
mcTb <- table(aq_ta$categories[smp], mcs$Prediction[smp])

# relative frequencies of matches
ikPr <- sum(diag(ikTb))/length(aq_ta$categories)
fpPr <- sum(diag(fpTb))/length(aq_ta$categories)
rpPr <- sum(diag(rpTb))/length(aq_ta$categories)
mcPr <- sum(diag(mcTb))/length(aq_ta$categories)

# print in a table 
probcov <- c(ikPr, fpPr, rpPr, mcPr)
names(probcov) <- "Probability of coverage"
methods <- c("Kriging", "Fixed path", "Random path", "Multinomial")
comparison_methods <- data.frame(methods, probcov)
comparison_methods

# in order to establish which is the best predictor, one should perform these simulations more than once. At each time, another 100 observations must be randomly selected.
```

